---
title: "Portfolio Exercise 1, Methods 3, 2021, autumn semester"
author: 'Alina Kereszt'
date: "13.12.2021"
output: html_document
---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, tidy = TRUE, tidy.opts = list(width.cutoff = 60))
setwd("~/GitHub/Coding/Methods 3/week_02")
pacman::p_load(tidyverse, rsq, lme4, MuMIn)
```

# Assignment 1: Using mixed effects modelling to model hierarchical data
In this assignment we will be investigating the _politeness_ dataset of Winter and Grawunder (2012) and apply basic methods of multilevel modelling. 

## Dataset
The dataset has been shared on GitHub, so make sure that the csv-file is on your current path. Otherwise you can supply the full path.

```{r}
politeness <- read.csv('politeness.csv') ## read in data
```

# Exercises and objectives
The objectives of the exercises of this assignment are:  
1) Learning to recognize hierarchical structures within datasets and describing them  
2) Creating simple multilevel models and assessing their fitness  
3) Write up a report about the findings of the study  

REMEMBER: In your report, make sure to include code that can reproduce the answers requested in the exercises below  
REMEMBER: This assignment will be part of your final portfolio

## Exercise 1 - describing the dataset and making some initial plots

### 1) Describe the dataset, such that someone who happened upon this dataset could understand the variables and what they contain  
**i. Also consider whether any of the variables in _politeness_ should be encoded as factors or have the factor encoding removed. Hint: ```?factor```**

Winter and Grawunder (2012) studied the phonetic profile of Korean speakers in formal and informal scenarios.

  - subject: The ID of the experimental participant. Factor.
  
  - gender: Participant's gender. Factor.
  
  - scenario: Experimental condition in the within-subject experiment. Factor.
  
  - attitude: Is the scenario formal or informal. Factor.
  
  - f0mn: Mean frequency of participants' voice pitch during the scenario (higher value indicates higher pitch). Numeric.
  
  - total_duration: The length of the interaction within the scenario. Numeric.
  
  - hiss_count: Number of hissing intakes of breath during the scenario. Numeric.
    
```{r}
# Remove NAs for cleanliness and because some functions cannot handle them
politeness <- politeness %>% 
  na.exclude(f0mn)
# While in some cases it may be appropriate to replace NAs with the mean value of the variable, here it is NOT - the data points are not independent of each other!!!

# Ensure correct data format
politeness$subject <- as.factor(politeness$subject)
politeness$gender <- as.factor(politeness$gender)
politeness$scenario <- as.factor(politeness$scenario)
politeness$attitude <- as.factor(politeness$attitude)
politeness$total_duration <- as.numeric(politeness$total_duration)
politeness$f0mn <- as.numeric(politeness$f0mn)
politeness$hiss_count <- as.numeric(politeness$hiss_count)
```
    
### 2) Create a new data frame that just contains the subject _F1_ and run two linear models; one that expresses _f0mn_ as dependent on _scenario_ as an integer; and one that expresses _f0mn_ as dependent on _scenario_ encoded as a factor.
```{r}
F1 <- politeness %>% 
  filter(subject == "F1") # filter for subject _F1_

class(F1$scenario) #check data format of _scenario_ - "factor"
scenario_int <- as.integer(F1$scenario) #copy _scenario_ as integer

model_fac <- lm(f0mn ~ scenario, data = F1) #model where _scenario_ is factor
model_int <- lm(f0mn ~ scenario_int, data = F1) #model where _scenario_ is integer

model_fac$coefficients
model_int$coefficients
```

**i. Include the model matrices, $X$ from the General Linear Model, for these two models in your report and describe the different interpretations of _scenario_ that these entail.**
```{r}
X_fac <- model.matrix(model_fac, data = F1)
X_int <- model.matrix(model_int, data = F1)

X_fac
X_int
```

When _scenario_ is taken to be a factor, we suppose it to be categorical - the scenarios are discrete entities that don't lie on the same continuum. Therefore, a separate slope is calculated for each level.

When _scenario_ is taken to be an integer, we suppose it to be a continuous variable - the scenarios represent incrementally different values along the same continuum. Therefore, only a single slope is calculated to summarize the values _scenario_ takes on.

**ii. Which coding of _scenario_, as a factor or not, is more fitting?**

As the scenarios are relatively independent of each other and likely create 'sub-groups' in the data, treating them as factors seems most reasonable.
      
### 3) Make a plot that includes a subplot for each subject that has _scenario_ on the x-axis and _f0mn_ on the y-axis and where points are color-coded according to _attitude_.
```{r}
ggplot(data = politeness, aes(x = scenario, y = f0mn)) +
  geom_point(aes(colour = attitude)) +
  facet_wrap(~subject) +
  guides(color = guide_legend("Attitude")) +
  scale_color_manual(labels = c("Informal", "Polite"), 
                     values = c("cornflowerblue", "coral1")) +
  labs(title = "Mean frequency of pitch by participant in formal and informal scenarios", 
       x = "Scenario",
       y = "Mean frequency of pitch") +
  theme_minimal()
```

**i. Describe the differences between subjects**

Based on a visual inspection, the plots appear to indicate that individuals have a higher mean frequency of pitch in informal scenarios compared to polite ones, and that females in general have a higher mean frequency of pitch than males do.
    
## Exercise 2  - comparison of models
```{r, include = FALSE}
#mixed.model <- lmer(formula=..., data=...)
#example.formula <- formula(dep.variable ~ first.level.variable + (1 | second.level.variable))
```

### 1) Build four models and do some comparisons:

**i. a single level model that models _f0mn_ as dependent on _gender_;**
```{r}
m1 <- lm(f0mn ~ gender, data = politeness)
```

**ii. a two-level model that adds a second level on top of i. where unique intercepts are modeled for each _scenario_;**
```{r}
m2 <- lme4::lmer(f0mn ~ gender + (1 | scenario), data = politeness)
```

**iii. a two-level model that only has _subject_ as an intercept;**
```{r}
m3 <- lme4::lmer(f0mn ~ gender + (1 | subject), data = politeness)
```

**iv. a two-level model that models intercepts for both _scenario_ and _subject_.**
```{r}
m4 <- lme4::lmer(f0mn ~ gender + (1 | scenario) + (1 | subject), data = politeness)
```

**v. which of the models has the lowest residual standard deviation, also compare the Akaike Information Criterion `AIC`?**
```{r}
# Residual standard deviation of the models:
c(sigma(m1),
  sigma(m2),
  sigma(m3),
  sigma(m4)
)

# AIC of the models:
AIC(m1, m2, m3, m4)
```

Model _m4_ has both the lowest residual standard deviation (30.65803) and lowest AIC score (2092.482), although not much lower than the residual standard deviation and AIC score of model _m3_.

**vi. which of the second-level effects explains the most variance?**
```{r}
# The amount of variance explained by the model is represented by R^2, the coefficient of determination. 
rsq(m2)
rsq(m3)
```

Both models _m2_ and _m3_ have the same fixed effects (which thereby explain the same amount of variance). _Subject_ (the random intercept in model _m3_) appears to explain almost 10 times as much variance as _scenario_ (random intercept in model _m2_).
    
### 2) Why is our single-level model bad?

**i. create a new data frame that has three variables, _subject_, _gender_ and _f0mn_, where _f0mn_ is the average of all responses of each subject, i.e. averaging across _attitude_ and _scenario_.**
```{r}
baddata <- politeness %>% 
  group_by(subject, gender) %>% 
  summarize(f0mn = mean(f0mn))
```

**ii. build a single-level model that models _f0mn_ as dependent on _gender_ using this new dataset.**
```{r}
badmodel <- lm(f0mn ~ gender, data = baddata)
```

**iii. make Quantile-Quantile plots, comparing theoretical quantiles to the sample quantiles) using `qqnorm` and `qqline` for the new single-level model and compare it to the old single-level model (from 1).i). Which model's residuals ($\epsilon$) fulfill the assumptions of the General Linear Model better?)**
```{r}
qqnorm(residuals(m1))
qqline(residuals(m1))

qqnorm(residuals(badmodel))
qqline(residuals(badmodel))
```

The residuals for both appear approximately normally distributed, however, there is skewness present in the residuals for model _m1_ that is not present in the residuals for model _badmodel_ (perhaps because by taking the mean of the _f0mn_ values, less outliers or extreme values are taken into consideration).

**iv. Also make a quantile-quantile plot for the residuals of the  multilevel model with two intercepts. Does it look alright?**
```{r}
qqnorm(residuals(m4))
qqline(residuals(m4))
```
The residuals once again appear approximately normally distributed, albeit with some positive kurtosis (values concentrated around the mean to a greater extent than in a classical normal distribution). The data also shows skewness at the extremes.

### 3) Plotting the two-intercepts model

**i. Create a plot for each subject, (similar to part 3 in Exercise 1), this time also indicating the fitted value for each of the subjects for each for the scenarios (hint use `fixef` to get the "grand effects" for each gender and `ranef` to get the subject- and scenario-specific effects)**
```{r, include = FALSE}
# The fitted values for a model can be found using the following function:
fitted(m4)

# Alternatively, one might first take the intercept. Then, by adding or subtracting the applicable fixed and random effects, the fitted value for each participant in each scenario may be found.
fixef(m4)
ranef(m4)

# E.g.: F1 is female. To find the fitted value for her pitch in the 1st scenario, one would take:
fittedF1 <- fixef(m4)[1] + ranef(m4)$subject[,1][1] + ranef(m4)$scenario[,1][1]
fittedF1
```
```{r}
#Plotted:
ggplot(data = politeness, aes(x = scenario, y = f0mn, color = attitude, group = subject)) +
  geom_point() +
  geom_point(aes(x = scenario, y = fitted(m4)), color = "black") +
  facet_wrap(~subject) +
  guides(color = guide_legend("Attitude")) +
  scale_color_manual(labels = c("Informal", "Polite", "Fitted values"), 
                     values = c("cornflowerblue", "coral1", "black")) +
  labs(title = "Fitted values (black) for mean frequency of pitch by participant in formal and informal scenarios", 
       x = "Scenario",
       y = "Mean frequency of pitch") +
  theme_minimal()
```
    
## Exercise 3 - now with attitude

### 1) Carry on with the model with the two unique intercepts fitted: _scenario_ and _subject_.
**i. now build a model that has _attitude_ as a main effect besides _gender_.**
```{r}
m5 <- lme4::lmer(f0mn ~ gender + attitude + (1 | scenario) + (1 | subject), data = politeness)
```

**ii. make a separate model that, besides the main effects of _attitude_ and _gender_, also includes their interaction**
```{r}
m6 <- lme4::lmer(f0mn ~ gender * attitude + (1 | scenario) + (1 | subject), data = politeness)
```

**iii. describe what the interaction term in the model says about Korean men's pitch when they are polite relative to Korean women's pitch when they are polite (you don't have to judge whether it is interesting).**
```{r, include = FALSE}
fixef(m6)
```

The interaction appears to show that much like that of women, Korean mens' pitch decreases in polite situations, albeit to a somewhat lesser extent than that of women, as evidenced by the term 'genderM:attitudepol' being positive.

### 2) Compare the three models (1. gender as a main effect; 2. gender and attitude as main effects; 3. gender and attitude as main effects and the interaction between them). For all three models model unique intercepts for _subject_ and _scenario_ using residual variance, residual standard deviation and AIC.
```{r}
# Residual variance:
c(sum(residuals(m4)^2),
  sum(residuals(m5)^2),
  sum(residuals(m6)^2)
)

# Residual standard deviation
c(sigma(m4),
  sigma(m5),
  sigma(m6)
)

# AIC scores
AIC(m4, m5, m6)
```

Model _m5_ has the lowest residual standard deviation, although the difference from the residual standard deviation of _m6_ is negligible. Model _m6_ has the lowest residual standard deviation and AIC score, with the scores of model _m5_ being almost as low.

Considering that these measures of model comparison yield similar results for both models _m5_ and _m6_, we decided to choose model _m5_ for our analysis, as we concluded that the added complexity of model _m6_ did not constitute a significant improvement and may potentially involve overfitting.

### 3)  Choose the model that you think describe the data the best - and write a short report on the main findings based on this model. At least include the following:
  i. describe what the dataset consists of  
  ii. what can you conclude about the effect of gender and attitude on pitch (if anything)?  
  iii. motivate why you would include separate intercepts for subjects and scenarios (if you think they should be included)  
  iv. describe the variance components of the second level (if any)  
  v. include a Quantile-Quantile plot of your chosen model
  
## REPORT

Winter and Grawunder (2012) studied the phonetic profile of Korean speakers in formal and informal scenarios in an experiment of both within- and between-participant design. The collected data contains the following variables:
    - subject: The ID of the experimental participant. Factor.
    - gender: Participant's gender. Factor.
    - scenario: Experimental condition in the within-subject experiment. Factor.
    - attitude: Is the scenario formal or informal. Factor.
    - f0mn: Mean frequency of participants' voice pitch during the scenario (higher value indicates higher pitch). Numeric.
    - total_duration: The length of the interaction within the scenario. Numeric.
    - hiss_count: Number of hissing intakes of breath during the scenario. Numeric.
    
We investigated the effects of gender and scenario on participant's pitch using the following linear mixed-effects model:

_m <- lmer(f0mn ~ gender + attitude + (1 | scenario) + (1 | subject), data = politeness)_

We decided to include separate intercepts for both _scenario_ and _subject_ as not doing so would have violated the assumption of independence for linear models: the structure of the data is hierarchical, and both the experimental condition and the individual participant introduce underlying categorization into the data.

We also decided to omit including an interaction between gender and scenario as the added complexity did not result in a significantly better fit.
```{r, echo = FALSE}
summary(m5)$coefficients
```

The output shows that pitch is significantly modulated by gender and attitude, p < .001, with women using a higher pitch than men, and informal scenarios eliciting a higher pitch than polite ones. The second-level variables had a variance of 584.4 (subject) and 106.4 (scenario), compared to the residual variance of 882.7.

The residuals of the model are normally distributed:
```{r, echo = FALSE}
qqnorm(residuals(m5))
qqline(residuals(m5))
```